Viscous Transport in Eroding Porous Media

### 2019-05-10 docs/transport/erosion_transport.tex
worked on numerics

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{s:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We are interested in computing the velocity, shear stress, vorticity,
and pressure of the Stokes double-layer potential.  We have shown that
all these quantities can be written in terms of Cauchy integrals and its
first two derivatives.  Therefore, we require quadrature methods to
evaluate the expressions in equation~\eqref{eqn:cauchy}.  The trapezoid
rule is appropriate if the integrands are sufficiently smooth, where the
smoothness depends on the number of discretization points.  However,
given a fixed resolution, if the target point $\xx$ is too close to $\bd
\Omega$, the trapezoid rule becomes inaccurate and the overall method is
unstable.  There have been a variety of near-singular integration
schemes developed, and we extend a method first outlined by Barnett et
al.~\cite{bar-wu-vee2015} which we will call the {\em Barycentric
quadrature rule}.  Here, we briefly describe this quadrature method, and
then derive a necessary extension to simulate erosion.

The Barycentric quadrature rule starts with a general quadrature rule
for $v(x)$
\begin{align}
  v(x) \approx \frac{1}{2\pi i} \sum_{j=1}^N 
    w_j \frac{\tau(y_j)}{x - y_j},
\end{align}
where $w_j$ are the quadrature weights and $y_j \in \Gamma$ are the
quadrature nodes.  We will use the trapezoid rule because of its
superior accuracy when $x$ is sufficiently far from $\Gamma$.  However,
for a fixed resolution $N$, the error grows without bound as $x$
approaches $\Gamma$.  As outline by Barnett et
al.~\cite{bar-wu-vee2015}, the error is uniformly bounded if the
Barycentric quadrature rule
\begin{align}
  v(x) \approx \frac{\sum\limits_{j=1}^N \frac{v_j^-}{y_j-x}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}
  \quad \text{and} \quad
  v'(x) \approx \frac{\sum\limits_{j=1}^N 
    \frac{v_j^- - v(x)}{(y_j - x)^2}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, 
\end{align}
is used, where $x$ is a target point that is interior to $\Gamma$,
and $v_j^-$ is the value of $v(y_j)$.  Modifications of this formula
exist if $x$ is exterior to $\Gamma$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature for $v''(x)$}
While accurate quadrature rules for $v(x)$ and $v'(x)$ have already been
developed~\cite{bar-wu-vee2015}, this work requires a similar quadrature
rule for $v''(x)$.  We mimic the approach developed
in~\cite{bar-wu-vee2015} by starting with the Cauchy integral identities
{\color{red}
\begin{alignat}{3}
1 &= \frac{1}{2\pi i}\int_{\Gamma} 
    \frac{1}{y-x} dy, &&x \in \Omega, \label{eqn:cauchyId1}\\
\frac{1}{x-a} &= \frac{-1}{2\pi i}\int_{\Gamma} 
    \frac{(y-a)^{-1}}{y-x} dy,  \quad &&x \in \Omega^c\label{eqn:cauchyId2}. 
\end{alignat}
and the contour integral 
\begin{align}\label{eqn:cauchyId3}
0=\oint_{\Gamma} \frac{1}{(y-x)^n}dy, \quad n \neq 1.
\end{align}
Then we apply equations~\eqref{eqn:cauchyId1} and~\eqref{eqn:cauchyId3} on
the second derivative of Cauchy integral in \eqref{eqn:cauchy} and get
\begin{align*}
0 &= \frac{1}{2\pi i} \int_{\Gamma}
    \frac{2v^-(y)}{(y - x)^3} \, dy -v''(x)\\
&= \frac{1}{2\pi i} \int_{\Gamma}
    \frac{2v^-(y)}{(y - x)^3} \, dy 
-v''(x)\frac{1}{2\pi i}\int_{\Gamma} 
    \frac{1}{y-x} dy\\
&= \frac{1}{2\pi i} \int_{\Gamma}
    \frac{2v^-(y)}{(y - x)^3} \, dy 
-\frac{1}{2\pi i}\int_{\Gamma} 
    \frac{(y-x)^2v''(x)}{(y-x)^3} dy
-\frac{1}{2\pi i}\int_{\Gamma} \frac{2v(x)}{(y-x)^3}dy
-\frac{1}{2\pi i}\int_{\Gamma} \frac{2(y-x)v'(x)}{(y-x)^3}dy\\
&= \frac{1}{2\pi i}\int_{\Gamma} 
    \frac{2v^-({ y})-2v(x)-2(y-x)v'(x)-(y-x)^2v''(x)}{(y-x)^3} dy.
\end{align*}
}
\todo[inline]{S-H: How do we get these identities?}
\begin{alignat}{3}
  0 &= \frac{1}{2\pi i}\int_{\Gamma} 
    \frac{2v^-({ y})-2v(x)-2(y-x)v'(x)-(y-x)^2v''(x)}{(y-x)^3} dy, &&x \in \Omega, \\
  0 &= \frac{1}{2\pi i}\int_{\Gamma}
    \frac{2v^+(y)-2v(x)-2(y-x)v'(x)-(y-a)^{-1}(x-a)(y-x)^2 v''(x)}
    {(y-x)^3} dy, \quad &&x \in \Omega^c.  \label{eqn:Identity}
\end{alignat}
{\color{red}
and the quadrature formulations are
\begin{alignat}{3}
  v''(x) &\approx
    \frac{2\sum\limits_{j=1}^N \frac{v_j^- -v(x)-(y_j-x)v'(x)}{(y_j-x)^3}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, &&x \in \Omega, \\
  v''(x) &\approx \frac{1}{x-a}\frac{2\sum\limits_{j=1}^N
    \frac{v_j^+ -v(x)-(y_j-x)v'(x)}{(y_j-x)^3}}{\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j}
    , \quad &&x \in \Omega^c.  \label{eqn:SecondBary}
\end{alignat}
where $a$ is a point inside the $\Omega$ in equations~\eqref{eqn:Identity} and \eqref{eqn:SecondBary}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fast Summation Methods}
\label{sec:fmm}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods}
\label{s:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have shown that the velocity, shear stress, vorticity, and pressure
of the double-layer potential can all be written in terms of Cauchy
integrals and its first two derivatives.  Therefore, we require
quadrature methods to evaluate the expressions in
equation~\eqref{eqn:cauchy}.  Since the integrands are all smooth and
periodic, the Trapezoid rule has spectral accuracy.  However, the
integrand is nearly singular if $x$ is close to $\bd\Omega$, and the
quadrature error is arbitrarily large if the resolution $N$ is fixed.
There have been a variety of near-singular integration schemes
developed, and we extend a method first outlined by Barnett et
al.~\cite{bar-wu-vee2015} which we call the {\em Barycentric quadrature
rule}.  This method first uses the density function $\tau(y)$ to compute
the boundary data of $v$, denoted by $v^-{y}$.  Then, it makes use
of the identity
\begin{align}
  v(x) = \int_{\Gamma} \frac{v^{-}(y)}{y-x} \, dy, 
  \quad x \in \Omega.
\end{align}

Here, we briefly describe this quadrature method, and then derive a
necessary extension to simulate erosion.  The Barycentric quadrature
rule starts with a quadrature rule with nodes $y_j \in \Gamma$ and
weights $w_j$. We use the trapezoid rule because of its superior
accuracy when $x$ is sufficiently far from $\Gamma$.  However, as $x$
approaches $\Gamma$, the integrand becomes nearly singular and the
quality of the trapezoid rule deteriorates. To eliminate the
nearly-singular integrand, we use the identity
\begin{align}
  \int_{\Gamma} \frac{v^{-}(y) - v(x)}{y - x} \,dy = 0.
\end{align}
Then, even if $x$ is close to $\Gamma$, the integrand is no longer
nearly-singular and the accuracy of the trapezoid rule is bounded
independent of $x$.  A similar identity identity involving $v'(x)$ also
holds.   Now, applying the spectrally accurate trapezoid rule, the
quadrature error of the Barycentric quadrature rule
\begin{align}
  v(x) \approx \frac{\sum\limits_{j=1}^N \frac{v_j^-}{y_j-x}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}
  \quad \text{and} \quad
  v'(x) \approx \frac{\sum\limits_{j=1}^N 
    \frac{v_j^- - v(x)}{(y_j - x)^2}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, 
  \label{eqn:vvpQuadrature}
\end{align}
is uniformly bounded~\cite{bar-wu-vee2015}.  A slight modification of
these formulae are applied if $x \in \Omega^{c}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Barycentric Quadrature for $v''(x)$}
While accurate quadrature rules for $v(x)$ and $v'(x)$ have already been
developed~\cite{bar-wu-vee2015}, this work requires a similar quadrature
rule for $v''(x)$.  We mimic the approach developed
in~\cite{bar-wu-vee2015}.  First, if $x \in \Omega$ and $n \in
\mathbb{N}$, then 
\begin{align}
  \frac{1}{2\pi i}\int_{\Gamma} \frac{1}{(y-x)^n}\, dy = 
  \left\{
    \begin{array}{ll}
      1, & n = 1, \\
      0, & n > 1. \\
    \end{array}
  \right.
\end{align}
Applying this identity to the Cauchy integral theorem for $v''(x)$, we
have
\begin{align}
  0 &= \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - v''(x) \\
    &= \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v^{-}(y)}{(y-x)^3}\,dy - 
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{(y-x)^2 v''(x)}{(y-x)^3}\,dy -
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2v(x)}{(y-x)^3}\, dy -
    \frac{1}{2\pi i} \int_{\Gamma} 
      \frac{2(y-x)v'(x)}{(y-x)^3}\, dy \\
    &= \frac{1}{2\pi i}
      \frac{2v^{-}(y) - v''(x)(y-x)^2 - 2v(x) - 2(y-x)v'(x)}
      {(y-x)^3}\, dy.
\end{align}
Since we have constructed the integrand so that it is bounded for for
all $x \in \Omega$, the quadrature error of the trapezoid rule is
acceptable.  Applying the quadrature rule and solving for $v''(x)$, the
Barycentric quadrature rule for the second derivative is
\begin{align}
  v''(x) &\approx \frac{2\sum\limits_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j}
    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}.
\end{align}
To compute $v(x)$ and $v'(x)$, the Barycentric quadrature rule in
equation~\eqref{eqn:vvpQuadrature} is used. If $x \in \Omega^c$, the
identity
\begin{align}
\frac{1}{x-a} &= -\frac{1}{2\pi i}\int_{\Gamma} 
    \frac{(y-a)^{-1}}{y-x} dy, 
\end{align}
is used, and the resulting Barycentric quadrature formula is
\begin{align}
  v''(x) &\approx \frac{1}{x-a}\frac{2\sum\limits_{j=1}^N
    \frac{v^{+}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}}
    {\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fast Summation Methods}
\label{sec:fmm}
The Barycentric quadrature rule achieves acceptable accuracy for
computing the velocity, shear stress, and vorticity.  However, applied
naively, it requires $\mathcal{O}(N^2)$ operations to directly.  Since
the shear stress must only be computed once per time step, and the
vorticity is only used as a post-processing step, we accept the
$\mathcal{O}(N^2)$ algorithm.  However, the velocity must be computed at
every GMRES iteration, so a fast algorithm is used.

To compute the velocity with $\mathcal{O}(N)$ operations, we start by
using the trapezoid rule accelerated with the fast multipole method
(FMM). Since the trapezoid rule is inaccurate when $x$ is close to
$\Gamma$, we replace the trapezoid rule with the Barycentric quadrature
rule for such points.

This
cost can be reduced to $\mathcal{O}(N)$ by applying the fast multipole
method (FMM).  Rather than developing an FMM for layer potentials that
involve high-order derivatives of the Green's functions, we

@@

%{\color{red}
%\begin{alignat}{3}
%1 &= \frac{1}{2\pi i}\int_{\Gamma} 
%    \frac{1}{y-x} dy, &&x \in \Omega, \label{eqn:cauchyId1}\\
%\frac{1}{x-a} &= \frac{-1}{2\pi i}\int_{\Gamma} 
%    \frac{(y-a)^{-1}}{y-x} dy,  \quad &&x \in \Omega^c\label{eqn:cauchyId2}. 
%\end{alignat}
%and the contour integral 
%\begin{align}\label{eqn:cauchyId3}
%0=\oint_{\Gamma} \frac{1}{(y-x)^n}dy, \quad n \neq 1.
%\end{align}
%Then we apply equations~\eqref{eqn:cauchyId1} and~\eqref{eqn:cauchyId3} on
%the second derivative of Cauchy integral in \eqref{eqn:cauchy} and get
%\begin{align*}
%0 &= \frac{1}{2\pi i} \int_{\Gamma}
%    \frac{2v^-(y)}{(y - x)^3} \, dy -v''(x)\\
%&= \frac{1}{2\pi i} \int_{\Gamma}
%    \frac{2v^-(y)}{(y - x)^3} \, dy 
%-v''(x)\frac{1}{2\pi i}\int_{\Gamma} 
%    \frac{1}{y-x} dy\\
%&= \frac{1}{2\pi i} \int_{\Gamma}
%    \frac{2v^-(y)}{(y - x)^3} \, dy 
%-\frac{1}{2\pi i}\int_{\Gamma} 
%    \frac{(y-x)^2v''(x)}{(y-x)^3} dy
%-\frac{1}{2\pi i}\int_{\Gamma} \frac{2v(x)}{(y-x)^3}dy
%-\frac{1}{2\pi i}\int_{\Gamma} \frac{2(y-x)v'(x)}{(y-x)^3}dy\\
%&= \frac{1}{2\pi i}\int_{\Gamma} 
%    \frac{2v^-({ y})-2v(x)-2(y-x)v'(x)-(y-x)^2v''(x)}{(y-x)^3} dy.
%\end{align*}
%}
%\todo[inline]{S-H: How do we get these identities?}
%\begin{alignat}{3}
%  0 &= \frac{1}{2\pi i}\int_{\Gamma} 
%    \frac{2v^-({ y})-2v(x)-2(y-x)v'(x)-(y-x)^2v''(x)}{(y-x)^3} dy, &&x \in \Omega, \\
%  0 &= \frac{1}{2\pi i}\int_{\Gamma}
%    \frac{2v^+(y)-2v(x)-2(y-x)v'(x)-(y-a)^{-1}(x-a)(y-x)^2 v''(x)}
%    {(y-x)^3} dy, \quad &&x \in \Omega^c.  \label{eqn:Identity}
%\end{alignat}
%{\color{red}
%and the quadrature formulations are
%\begin{alignat}{3}
%  v''(x) &\approx
%    \frac{2\sum\limits_{j=1}^N \frac{v_j^- -v(x)-(y_j-x)v'(x)}{(y_j-x)^3}w_j}
%    {\sum\limits_{j=1}^N \frac{1}{y_j-x}w_j}, &&x \in \Omega, \\
%  v''(x) &\approx \frac{1}{x-a}\frac{2\sum\limits_{j=1}^N
%    \frac{v_j^+ -v(x)-(y_j-x)v'(x)}{(y_j-x)^3}}{\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j-x}w_j}
%    , \quad &&x \in \Omega^c.  \label{eqn:SecondBary}
%\end{alignat}
%where $a$ is a point inside the $\Omega$ in equations~\eqref{eqn:Identity} and \eqref{eqn:SecondBary}.
%
%}
